A Model for Extracting Time Delays from Noisy Time Series Data

### 2013-08-23 doc/Paper1/ms.tex
Filling in parameter table

\documentclass[useAMS,usenatbib, a4paper]{mn2e} \usepackage{natbib} 
\addtolength{\voffset}{-0.5in}
\usepackage{graphicx} \usepackage{amssymb}
\usepackage{amsmath}

@@

An introduction to Gaussian Processes and their applications can be found in
\citet{rasmussen}. In their applications, GPs are usually used as a choice of
prior distribution over the space of functions. In this work, GPs describe the
sampling distributions, or the probability of the data given the model
parameters. This is a common situation in astrophysics in fields where the
signals are complex and the best models we can think of are 
stochastic processes. Two prominent examples are quasar
variability \citep{2009ApJ...698..895K, 2010ApJ...721.1014M, 2011ApJ...735...80Z} and solar-like oscillations of stars \citep{2009MNRAS.395.2226B}.

@@

\section{Bayesian Inference}
For inferring unknown parameters $\params$ from data $\data$ in the context
of prior information $\info$,
Bayes' rule gives the posterior distribution
\begin{eqnarray}
p(\params | \data, \info) \propto p(\params|\info)p(\data|\params, \info)
\end{eqnarray}
The prior distribution $p(\params|\info)$ describes prior knowledge or uncertainty
about the values of the parameter(s). The ``sampling distributions''
$p(\data|\params,\info)$
describes prior knowledge about the data we expect to observe, as a function of
$\params$ (Caticha). When $\data$ is fixed $p(\data|\params,\info)$ is a function of
$\params$ only and is called the likelihood function.

We begin by defining our model for the prior knowledge about how the time
delays give rise rise to the observed data. This will determine the sampling
distribution, and also the set of parameters (including nuisance parameters) to
be inferred.

\section{Sampling Distributions}
The intrinsic quasar variability, in magnitudes, and in whatever band the
observations are carried out in, can be described by a function $q(t)$. For a
two-image system, the resulting light curves, if observed without gaps and
noise, would be described by the functions
\begin{eqnarray}
y_1(t) &=& m_1 + f(t) + \mu_1(t) \\
y_2(t) &=& m_2 + f(t - \tau_2) + \mu_2(t) \\
\end{eqnarray}
where $\{m_1, m_2\}$ are the mean magnitudes of the images,
$\tau_2$ is the time delay of image 2 relative to image 1, and the
functions $\{\mu_1(t), \mu_2(t)\}$ describe any perturbations that are not
shared by both images. A major source of perturbations is the effect of
microlensing on each of the light curves. The data then consist of
noisy, irregularly sampled measurements of the light curves $\{y_1(t), y_2(t)\}$.

@@

\begin{table*}
\begin{tabular}{lll}
\hline
{\bf Parameter} &	{\bf Meaning}			& {\bf Prior}\\
\hline
$\tau_2$	&	Time delay relative to image 1	& Cauchy()$T(,)$\\
$\{m_1, m_2\}$	&	Mean magnitude for each image	& Uniform()\\
\hline
{\bf Data}	&	{\bf Meaning}			& {\bf Sampling Distribution}\\
\hline
$\{Y_1, ..., Y_N\}$ &	Noisy magnitude measurements	& $\mathcal{N}\left(\mathbf{m}(\params), \mathbf{C}(\params)\right)$\\
\hline
{\bf Prior Information} & {\bf Meaning}			&\\
\hline
$N$		&	Number of measurements (all images together)\\
$\{t_1, ..., t_N\}$ &	Measurement times\\
$\{k_1, ..., k_N\}$ &	Flags identifying which image was observed
\end{tabular}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\begin{center}
\begin{tabular}{ccc}
Parameter & \vline & Prior \\
\hline
$\tau$ & \vline & $U(-\Delta t/2, \Delta t/2)$ \\ 
$m$ & \vline & $U(y_{\rm min}, y_{\rm max})$ \\
$\sigma_{\rm QSO}$ & \vline & $LU(10^{-3}, 10)$ \\
$L_{\rm QSO}$ & \vline & $LU(10^{-3}\Delta t, 10\Delta t)$ \\
$\sigma_{\rm micro}$ & \vline & $LU(10^{-3}, 10)$ \\
$L_{\rm micro}$ & \vline & $LU(10^{-3}\Delta t, 10\Delta t)$ \\
$\alpha_{\rm micro}$ & \vline & $U(1, 2)$ \\
$\kappa$ & \vline & $LU(1, 10)$
\end{tabular}
\end{center}
\caption{Prior probability distributions for the parameters. $U(a, b)$ denotes
a uniform prior between $a$ and $b$, and $LU(a, b)$ denotes a log-uniform
(that is, uniform in the logarithm of the parameter) prior $\propto 1/x$
between $a$ and $b$. $\Delta t$ is the duration of the data set. All priors
are independent.\label{priors}}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

>>>>

\documentclass[useAMS,usenatbib, a4paper]{mn2e} \usepackage{natbib} 
\addtolength{\voffset}{-0.5in}
\usepackage{graphicx} \usepackage{amssymb}
\usepackage{amsmath}
\usepackage{dsfont}

@@

Part of our model involves Gaussian Processes, which are a family of probability
distributions over the space of possible functions.
An introduction to Gaussian Processes and their applications can be found in
\citet{rasmussen}. In their applications, GPs are usually used as a choice of
prior distribution over the space of functions. In this work, GPs describe the
sampling distributions, or the probability of the data given the model
parameters. This is a common situation in astrophysics in fields where the
signals are complex and the best models we can think of are 
stochastic processes. Two prominent examples are quasar
variability \citep{2009ApJ...698..895K, 2010ApJ...721.1014M, 2011ApJ...735...80Z}
and solar-like oscillations of stars \citep{2009MNRAS.395.2226B}.

In Section~\ref{sec:bayes} we briefly discuss Bayesian Inference, and in
Section~\ref{sec:sampling} we describe our specific
modelling assumptions. The computational implementation is described in
Section~\ref{sec:computation}. In Section~\ref{sec:tests} we test the inference
with two simulated data sets, one ``easy'' data set and one ``hard'' data set.
In Section~\ref{sec:b1608} we apply the model to B1608+656.

@@

\section{Bayesian Inference}\label{sec:bayes}
For inferring unknown parameters $\params$ from data $\data$ in the context
of prior information $\info$,
Bayes' rule gives the posterior distribution
\begin{eqnarray}
p(\params | \data, \info) \propto p(\params|\info)p(\data|\params, \info)
\end{eqnarray}
The prior distribution $p(\params|\info)$ describes prior knowledge or uncertainty
about the values of the parameter(s). The ``sampling distributions''
$p(\data|\params,\info)$
describe prior knowledge about the data we expect to observe, as a function of
$\params$ \citep{caticha}. When $\data$ is fixed $p(\data|\params,\info)$ is a function of
$\params$ only and is called the likelihood function.

We begin by defining our model for the prior knowledge about how the time
delays give rise rise to the observed data. This will determine the sampling
distributions, and also the set of parameters (including nuisance parameters) to
be inferred. Our choice of prior distribution for the parameters will be
discussed in Section

\section{Sampling Distributions}\label{sec:sampling}
The intrinsic quasar variability, in magnitudes, and in whatever band the
observations are carried out in, can be described by a function $q(t)$. For a
two-image system, the resulting light curves, if observed without gaps and
noise, would be described by the functions
\begin{eqnarray}
y_1(t) &=& m_1 + f(t) + \mu_1(t) \\
y_2(t) &=& m_2 + f(t - \tau_2) + \mu_2(t) \\
\end{eqnarray}
where $\{m_1, m_2\}$ are the mean magnitudes of the images,
$\tau_2$ is the time delay of image 2 relative to image 1, and the
functions $\{\mu_1(t), \mu_2(t)\}$ describe any perturbations that are not
shared by both images. A major source of perturbations is the effect of
microlensing on each of the light curves. The data then consist of
noisy, irregularly sampled measurements of the light curves
$\{y_1(t), y_2(t)\}$.

@@

\begin{table*}
\begin{tabular}{lll}
\hline
{\bf Parameter} &	{\bf Meaning}			& {\bf Prior}\\
\hline
$\tau_2$	&	Time delay relative to image 1	& Cauchy$\left(0, 0.1\tau_{\rm range}\right)T\left(\tau_{\rm min},\tau_{\rm max}\right)$\\
$\{m_1, m_2\}$	&	Mean magnitude for each image	& Uniform$(m_{\rm min}, m_{\rm max})$\\
$L$		&	Length scale of QSO variations	& LogUniform$(L_{\rm min}, L_{\rm max})$\\
$\beta$		&	Size of QSO variations		& LogUniform$(\beta_{\rm min}, \beta_{\rm max})$\\
$A$		&	Size of image-specific variations	& LogUniform$\left(A_{\rm min}, A_{\rm max}\right)$\\
$\{\ell_1, \ell_2\}$ &	Length scale of image-specific variations & LogUniform$(\ell_{\rm min}, \ell_{\rm max})$\\
$\alpha$	&	Smoothness of image-specific variations	& Uniform$(1, 2)$\\
$f_{\rm bad}$	&	Fraction of ``bad'' data points	& Uniform$(0,1)$\\
$\kappa_1$	&	Error-bar boost for ``good'' data & $\frac{1}{2}\delta(\kappa_1 - 1) + \frac{1}{2}$LogUniform$(1,100)$\\
$\kappa_2$	&	Error-bar boost for ``bad'' data & LogUniform$(\kappa_1, 100\kappa_1)$\\
$\{b_i\}_{i=1}^N$ &	Badness hidden variables for each data point & Uniform$(0,1)$\\
\hline
{\bf Derived Parameter} & {\bf Meaning}	&	{\bf Definition}\\
\hline
$\{B_i\}_{i=1}^N$	&	Badness flag for each data point & $\mathds{1}\left(b_i \leq f_{\rm bad}\right)$\\
\hline
{\bf Data}	&	{\bf Meaning}			& {\bf Sampling Distribution}\\
\hline
$\{Y_1, ..., Y_N\}$ &	Noisy magnitude measurements	& $\mathcal{N}\left(\mathbf{m}(\params), \mathbf{C}(\params)\right)$\\
\hline
{\bf Prior Information} & {\bf Meaning}			&\\
\hline
$N$		&	Number of measurements (all images together)\\
$\{t_1, ..., t_N\}$ &	Measurement times\\
$\{k_1, ..., k_N\}$ &	Flags identifying which image was observed
\end{tabular}
\end{table*}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
\begin{center}
\begin{tabular}{ccc}
Parameter & \vline & Prior \\
\hline
$\tau$ & \vline & $U(-\Delta t/2, \Delta t/2)$ \\ 
$m$ & \vline & $U(y_{\rm min}, y_{\rm max})$ \\
$\sigma_{\rm QSO}$ & \vline & $LU(10^{-3}, 10)$ \\
$L_{\rm QSO}$ & \vline & $LU(10^{-3}\Delta t, 10\Delta t)$ \\
$\sigma_{\rm micro}$ & \vline & $LU(10^{-3}, 10)$ \\
$L_{\rm micro}$ & \vline & $LU(10^{-3}\Delta t, 10\Delta t)$ \\
$\alpha_{\rm micro}$ & \vline & $U(1, 2)$ \\
$\kappa$ & \vline & $LU(1, 10)$
\end{tabular}
\end{center}
\caption{Prior probability distributions for the parameters. $U(a, b)$ denotes
a uniform prior between $a$ and $b$, and $LU(a, b)$ denotes a log-uniform
(that is, uniform in the logarithm of the parameter) prior $\propto 1/x$
between $a$ and $b$. $\Delta t$ is the duration of the data set. All priors
are independent.\label{tab:prior}}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\bibitem[\protect\citeauthoryear{Caticha}{2008}]{caticha} 
Caticha A., 2008, arXiv, arXiv:0808.0012
